One page description

### 2016-12-08 publications/Journal/6-allocation.tex
Review section 6
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>
Made whole paper (more) consistent

Although the synchronization protocol is one of the defining factors in simulation performance, model allocation has a significant impact on which protocol is ideal.
Depending on the model structure, and how it is mapped to the different kernels, it might not even make sense to parallelize at all.
Indeed, if the model is allocated such that frequent communication is necessary between kernels, parallelism is naturally reduced.
This brings us to the topic of model allocation, as also implemented in \textit{PythonPDEVS}~\cite{PythonPDEVS2}.

The modeller can specify which kernel a model should be allocated to, should such manual intervention be required.
This is handled by the default model allocator.
If no preference is given, a simple striping scheme is used but this is often insufficient.
By overriding the default allocator, a modeller tunes the allocation scheme for a specific model, maximizing parallel speedup.
This interface can be linked to graph algorithms for automatic allocation scheme generation, for example to avoid cycles in the dependency graph.

@@

Unlike the Queue model, the width of the hierarchy is still present in the topology of the atomic models after direct connection.
The PHOLDTree model allows us to investigate parallel speedup in terms of model allocation, by modifying the depth and width (fanout) model parameters.

@@

PHoldtree, like Queue, is a highly hierarchical model but one where the flattened structure cannot be partitioned into a chain, as was the case in the Queue model.
This topology is interesting since it highlights the effects of allocation.
First, we evaluate the model in sequential simulation to provide a baseline for parallel simulation.

\subsubsection{Sequential Simulation}
Since \textit{adevs} does not use direct connection, we expect a noticeable performance difference between \textit{dxex} and \textit{adevs}.
This is shown in Figure~\ref{fig:PHOLDtree_seq_dn_benchmark}, where the fanout ($n$) determines the performance penalty \textit{adevs} suffers compared to \textit{dxex}.
Profiling indeed indicates that an increase in width per subtree ($n$) leads to higher overheads in \textit{adevs} due to the lack of direct connection.
\textit{Dxex} uses direct connection, making it independent of fanout.
Performance of \textit{dxex} is, in this model, only dependent on the number of models.
Slight deviations can still be seen, though, caused by the initialization overhead of direct connection.
Both \textit{adevs} and \textit{dxex} scale linearly in the number of atomic models.

\begin{figure}
    \center
    \includegraphics[width=\columnwidth]{fig/pholdtree_sequential_dn.eps}
    \caption{Effect of hierarchy in sequential simulation.}
    \label{fig:PHOLDtree_seq_dn_benchmark}
\end{figure}

\subsubsection{Parallel Simulation}
Next, we run the model using two different model allocation schemes: breadth-first and depth-first.
But first, we explain what we mean by both allocation schemes.

With breadth-first allocation, we traverse the tree in a breadth-first way, allocating subsequently visited atomic models to the same node.
This means, intuitively, that atomic models at the same level in the tree, but not necessarily siblings, are frequently allocated to the same node.
Since there is only infrequently some communication between siblings, and even never between different subtrees, this does not sound an intuitive allocation.
This allocation strategy is shown in Figure~\ref{fig:PholdTree_model_bfs}.

@@

With depth-first allocation, we traverse the tree in a depth-first way, allocating subsequently visited atomic models to the same node.
This means, intuitively, that subtreees are frequently allocated to the same node.
This allocation strategy is shown in Figure~\ref{fig:PholdTree_model_dfs}.

@@

Both allocators will try to divide models evenly over kernels.
The effects of varying the number of models per kernel are already evaluated in the previous section on scaling.
Here we want to highlight the overhead of communication and inter-kernel dependence.

The breadth first allocation scheme results in a dependency chain with multiple branches, much like in the Queue model.
Such a linear dependency chain can result in a parallel speedup as we demonstrated with the Queue model.
This is not always true though: a single kernel with an unbalanced number of atomic models or unequal computational load in transition functions slows down the remainder of the chain.
This effect is also apparent if the thread a kernel runs on is not fairly scheduled.
With conservative synchronization this leads to excessive polling of the EOT of the other kernels.
With optimistic synchronization this leads to a cascade of rollbacks, since dependent kernels will simulate ahead of the slower kernel.

After simulation the traces can be visualized for both breadth-first and depth-first allocation.
Using a breadth-first allocation scheme, as shown in Figure~\ref{fig:pholdtree_visualize_parBFS}, we notice that many events get exchanged between kernels.
This is caused by the high number of inter-kernel connections and the high number of events exchanged over these connections.
The number of connections between nodes at the same simulation kernel is also rather low.
Using a depth-first allocation scheme, however, as shown in Figure~\ref{fig:pholdtree_visualize_parDFS}, minimizes inter-kernel connections while maximizing intra-kernel connections.

@@

Simulation results are shown in Figure~\ref{fig:PholdTree_plot_alloc_high} for both allocation schemes in combination with both synchronization protocols.
We see that for both synchronization protocols, the depth-first allocation is significantly better than breadth-first allocation.
This is what we expected for this model: depth-first allocation maintains locality better than breadth-first allocation.
Whereas this is the case in this example, this is not true in general, as the ideal allocation depends on the model being simulated.

The most prominent aspect of these results is the low performance for conservative depth-first allocation for two kernels.
This is mostly caused by the difference between a sequential simulation and a parallel simulation: suddenly we need to take into account synchronization and passing around of lookahead values.
And since the number of kernels is low, the overhead dominates any further speedup.
Optimistic is less sensitive to the number of models per kernel as it does not need to poll each model for a lookahead, this explains the lower runtime penalty observed for optimistic.

%The effect is due to the topology of the kernels, in BFS the risk on reverts is higher. Optimistic is less influenced but will get a lower speedup in the good case.
Interestingly, we see that optimistic synchronization is less influenced by the allocation than conservative synchronization.
This is likely caused by the lower number of connections to take into account in conservative synchronization.
Whereas conservative synchronization needs to take into account even scarcely used connections, optimistic synchronization does not.
The same is true in the opposite direction, though, where optimistic synchronization is slower when a good allocation is chosen.
Conservative synchronization will then be able to make better estimates, whereas optimistic synchronization does not make estimations.

>>>>

Although the synchronization protocol is one of the defining factors in simulation performance, model allocation has a significant impact on which protocol is ideal.
Depending on the model structure, and how it is mapped to the different kernels, it might not even make sense to parallelize at all.
Indeed, if the model is allocated such that frequent communication is necessary between kernels, parallelism is naturally reduced.
This brings us to the topic of model allocation, as also implemented in \textit{PythonPDEVS}~\cite{PythonPDEVS2}.
In this section we focus on the inter kernel synchronization protocols, as intra kernel synchronization is unaffected by the inter kernel topology.

The modeller can specify which kernel a model should be allocated to, should such manual intervention be required.
This is handled by the default model allocator.
If no preference is given, a simple striping scheme is used but this is often insufficient.
By overriding the default allocator, a modeller tunes the allocation scheme for a specific model.
This interface can be linked to graph algorithms for automatic allocation scheme generation, for example to avoid cycles in the dependency graph.

@@

Unlike the Queue model, the width of the hierarchy is still present in the topology of the atomic models after direct connection.
The PHOLDTree model allows us to investigate speedup in terms of model allocation, by modifying the depth and width (fanout) model parameters.

@@

PHoldtree, like Queue, is highly hierarchical, but its flattened structure cannot be partitioned into a chain, as was the case in the Queue model.
This topology is interesting since it highlights the effects of allocation.
First, we evaluate the model with \textit{NN} synchronization to provide a baseline for further results.

\subsubsection{No Inter-Kernel Synchronization (\textit{NN})}
Since \textit{adevs} does not use direct connection, we expect a noticeable performance difference between \textit{dxex} and \textit{adevs}.
This is shown in Figure~\ref{fig:PHOLDtree_seq_dn_benchmark}, where the fanout ($n$) determines the performance penalty \textit{adevs} suffers compared to \textit{dxex}.
Profiling indeed indicates that an increase in width per subtree ($n$) leads to higher overheads in \textit{adevs} due to the lack of direct connection.
\textit{Dxex} uses direct connection, making it independent of fanout.
Performance of \textit{dxex} is, for this model, only dependent on the number of models.
Slight deviations can still be seen, though, caused by the initialization overhead of direct connection.
Both \textit{adevs} and \textit{dxex} scale linearly in the number of atomic models.

\begin{figure}
    \center
    \includegraphics[width=\columnwidth]{fig/pholdtree_sequential_dn.eps}
    \caption{Effect of hierarchy with \textit{NN} synchronization.}
    \label{fig:PHOLDtree_seq_dn_benchmark}
\end{figure}

\subsubsection{Inter-Kernel Synchronization (\textit{CN} and \textit{ON})}
Next, we run the model using two different allocation schemes: breadth-first and depth-first.
First, we explain both allocation schemes.

Breadth-first allocation traverses the tree in a breadth-first way, allocating subsequently visited atomic models to the same node.
Intuitively, atomic models at the same level in the tree, but not necessarily siblings, are frequently allocated to the same node.
Since there is only infrequently some communication between siblings, and even never between different subtrees, this does not sound an intuitive allocation.
This allocation strategy is shown in Figure~\ref{fig:PholdTree_model_bfs}.

@@

Depth-first allocation traverses the tree in a depth-first way, allocating subsequently visited atomic models to the same node.
Intuitively, subtrees are frequently allocated to the same node, as shown in Figure~\ref{fig:PholdTree_model_dfs}.

@@

Both allocators will try to divide models evenly over kernels.
The effects of varying the number of atomic models per kernel are already evaluated in the previous section on scaling.
Here we want to highlight the overhead of communication and inter-kernel dependence.

The breadth first allocation scheme results in a dependency chain with multiple branches, much like in the Queue model.
Such a linear dependency chain can result in a higher speedup, as demonstrated with the Queue model.
This is not always true though: a single kernel with an unbalanced computational load slows down the remainder of the chain.
This effect is also apparent if the thread a kernel runs on is not fairly scheduled by the operating system.
With conservative synchronization this leads to excessive polling of the EOT of the other kernels.
With optimistic synchronization this leads to a cascade of rollbacks, since dependent kernels will simulate ahead of the slower kernel.

After simulation the traces can be visualized for both breadth-first and depth-first allocation.
Using a breadth-first allocation scheme, as shown in Figure~\ref{fig:pholdtree_visualize_parBFS}, we notice that many events get exchanged between kernels.
This is caused by the high number of inter-kernel connections and the high number of events exchanged over these connections.
The number of connections between atomic models at the same simulation kernel is also rather low.
Using a depth-first allocation scheme, however, as shown in Figure~\ref{fig:pholdtree_visualize_parDFS}, minimizes inter-kernel connections while maximizing intra-kernel connections.

@@

Simulation results are shown in Figure~\ref{fig:PholdTree_plot_alloc_high} for both allocation schemes in combination with both synchronization protocols.
We see that for both synchronization protocols, the depth-first allocation is significantly better than breadth-first allocation.
This is what we expected for this model: depth-first allocation preserves locality better than breadth-first allocation.
Whereas this is the case in this example, this is not true in general, as the ideal allocation depends on the model being simulated.

The most prominent aspect of these results is the low performance for conservative depth-first allocation for two kernels.
This is mostly caused by the introduction of synchronization protocols: suddenly we need to take into account other kernels and passing around of lookahead values.
And since the number of kernels is low, the overhead dominates.
Optimistic is less sensitive to the number of atomic models per kernel as it does not need to poll each model for a lookahead, this explains the lower runtime penalty observed for optimistic.

Interestingly, we see that optimistic synchronization is less influenced by the allocation than conservative synchronization.
This is likely caused by the lower number of connections to take into account in conservative synchronization.
Whereas conservative synchronization needs to take into account even scarcely used connections, optimistic synchronization does not.
The same is true in the opposite direction, though, where optimistic synchronization is slower when a good allocation is chosen.
Conservative synchronization will then be able to make better estimates, whereas optimistic synchronization does not make estimations.

### 2016-12-08 publications/Journal/7-related-work.tex
Made whole paper (more) consistent

In terms of code design and philosophy, \textit{dxex} is most related to \textit{PythonPDEVS}~\cite{PythonPDEVS}.
Performance of \textit{PythonPDEVS} was still decent through the use of ``hints'' from the modeler.
In this spirit, we offer users the possibility to choose between different synchronization protocols.
This allows users to choose the most appropriate synchronization protocol, depending on the model.
Contrary to \textit{PythonPDEVS}, however, \textit{dxex} doesn't support distributed simulation~\cite{JDF}, model migrations~\cite{PythonPDEVS2}, or activity hints~\cite{PythonPDEVS_ACTIMS}.

@@

Previous \textsf{DEVS} simulation tools have already implemented multiple synchronization protocols, though none have done so in a strictly modular way that allows straightforward protocol switching for a single given model.
For example, \textit{adevs} only supports conservative synchronization, and \textit{vle} only supports experiment-level parallelism (\textit{i.e.}, run multiple experiments in parallel).
Closest to our support for multiple synchronization protocols is \textit{CD++}~\cite{CD++}.
For \textit{CD++}, both a conservative (\textit{CCD++}~\cite{CCD++}) and optimistic (\textit{PCD++}~\cite{PCD++}) variant exist.
Despite the implementation of both protocols, they are entirely different projects.
Some features might therefore be implemented in \textit{CCD++} and not in \textit{PCD++}, or vice versa.
And while this might not be a problem at this time, the problem will only get worse when each project follows its own course.
\textit{Dxex}, on the other hand, is a single project, where the choice of synchronization protocol is a simple configuration option.
\textit{CD++}, however, implements both conservative and optimistic synchronization for distributed simulation, whereas we limit ourselves to parallel simulation. 
A new architecture for sequential PDEVS simulation has been introduced in \cite{SeqPDEVSArch} with promising performance results. This algorithm achieves a speedup by using the inherent parallelism in PDEVS without the need for synchronization.
By limiting our approach to a pure parallel simulation, we are able to achieve higher speedups through the use of shared memory communication. 
Recent work on parallel speedup in DEVS investigates theoretical limits of PDEVS~\cite{activitypdevs, amdahlpdevs}

@@

Model allocation and its impact on parallel performance has previously also been studied in the \textsf{PDES} community~\cite{PDESpartitioning}.
Referenced as partitioning of the simulation model, most studies distinguish between communication and computation as the two dimensions to partition over.
Partitioning a simulation model is identified as an issue to achieve scalability~\cite{Scalability}. 
Some research in this context has been done for \textsf{Parallel DEVS}~\cite{PDEVSpartitioning, NonFragmenting}.
Our contribution studies the effect of partitioning with emphasis on the effect of communication between processes and in the presence of a flattened hierarchy. 
We focus on static partitioning since this is a limiting factor for our conservative synchronization implementation which does not support model migration.
Model migration, as implemented by \textit{PythonPDEVS}, might be an interesting addition to model allocation.

>>>>

In terms of code design and philosophy, \textit{dxex} is most related to \textit{PythonPDEVS}~\cite{PythonPDEVS}.
Performance of \textit{PythonPDEVS} was decent through the use of ``hints'' from the modeler.
In this spirit, we offer users the possibility to choose between different synchronization protocols.
This allows users to choose the most appropriate synchronization protocol, depending on the model.
Contrary to \textit{PythonPDEVS}, however, \textit{dxex} doesn't support distributed simulation~\cite{JDF}, model migrations~\cite{PythonPDEVS2}, or activity hints~\cite{PythonPDEVS_ACTIMS}.

@@

Previous \textsf{DEVS} simulation tools have already implemented multiple synchronization protocols, though none have done so in a strictly modular way that allows straightforward protocol switching for a single given model.
For example, \textit{adevs} only supports conservative synchronization, and \textit{vle} only supports experiment-level parallelism (\textit{i.e.}, run independent experiments in parallel).
Closest to our support for multiple synchronization protocols is \textit{CD++}~\cite{CD++}.
For \textit{CD++}, both a conservative (\textit{CCD++}~\cite{CCD++}) and optimistic (\textit{PCD++}~\cite{PCD++}) variant exist.
Despite the implementation of both protocols, they are entirely different projects.
Some features might therefore be implemented in \textit{CCD++} and not in \textit{PCD++}, or vice versa.
And while this might not be a problem at this time, the problem will only get worse when each project follows its own course.
\textit{Dxex}, on the other hand, is a single project, where the choice of synchronization protocol is a simple configuration option.
\textit{CD++}, however, implements both conservative and optimistic synchronization for distributed simulation, whereas we limit ourselves to parallel simulation. 
A new architecture for sequential PDEVS simulation has been introduced in \cite{SeqPDEVSArch} with promising performance results.
This algorithm achieves a speedup by using \textit{xP} synchronization.
By limiting our approach to pure parallel simulation (\textit{i.e.}, no distributed simulation), we are able to achieve higher speedups through the use of shared memory communication. 
Recent work on parallel speedup in DEVS investigates theoretical limits of PDEVS~\cite{activitypdevs, amdahlpdevs}

@@

Model allocation and its impact on parallel performance has previously also been studied in the \textsf{PDES} community~\cite{PDESpartitioning}.
Referenced as partitioning of the simulation model, most studies distinguish between communication and computation as the two dimensions to partition over.
Partitioning a model is identified as an issue to achieve scalability~\cite{Scalability}. 
Some research in this context has been done for \textsf{Parallel DEVS}~\cite{PDEVSpartitioning, NonFragmenting}.
Our contribution studies the effect of partitioning with emphasis on the effect of communication between kernels and in the presence of a flattened hierarchy. 
We focus on static partitioning since this is a limiting factor for our conservative synchronization implementation which does not support model migration.
Model migration, as implemented by \textit{PythonPDEVS}, might be an interesting addition to model allocation.

### 2016-12-08 publications/Journal/8-conclusions.tex
Small changes to section 8
Made whole paper (more) consistent

In this paper, we introduced \textit{DEVS-Ex-Machina} (``\textit{dxex}''), a new C++11-based \textsf{Parallel DEVS} simulation tool.
Our main contribution is the implementation of multiple synchronization protocols for parallel multicore simulation.
We have shown that there are indeed models which can be simulated significantly faster using either synchronization protocol.
\textit{Dxex} allows the user to choose between either conservative or optimistic synchronization at the start of simulation.
Depending on observed model behaviour and simulation performance, runtime switching between synchronization protocols can be used.

Notwithstanding our modularity, \textit{dxex} achieves performance competitive to \textit{adevs}, another very efficient \textsf{DEVS} simulation tool.
Performance is measured both in elapsed time, and memory usage.
Our empirical analysis shows that allocation of models over kernels is critical to enable a parallel speedup. Furthermore we have shown when and why optimistic synchronization can outperform conservative and vice versa. Finally we investigated the effect of memory (de)allocation on parallel simulation.

Future work is possible in several directions.
First, our implementation of optimistic synchronization should be more tolerant to low-memory situations.
In its current state, simulation will simply halt with an out-of-memory error. The use of transactional memory can offer several advantages in this project. If it becomes part of the new C++17 standard it would be of great interest to see if it can help reduce the performance effects of memory allocation and synchronization.
Having simulation control, which can throttle the speed of nodes that use up too much memory, has been shown to work in these situations~\cite{FujimotoBook}.
Faster GVT implementations~\cite{Fujimoto:1997:CGV:268403.268404,Bauer:2005:SND:1069810.1070159} might further help to minimize this problem.
Second, the runtime switching between synchronization protocols can be driven using machine learning techniques.
The simulation engine is already capable of collecting data to inform such a process, and is designed to listen for commands from an external component.
Third, automatic allocation might be possible by analysis of the connections between models.
This information is already used in \textit{dxex} to construct the dependency graph in conservative synchronization.
A graph algorithm that distributes models, while avoiding cycles, could be used to offer a parallel speedup in either optimistic or conservative synchronization.
Similarly, it could serve as a default parallel allocation scheme that can be improved by the user.

>>>>

In this paper, we introduced \textit{DEVS-Ex-Machina} (``\textit{dxex}''), a new C++11-based \textsf{Parallel DEVS} simulation tool.
Our main contribution is the implementation of multiple synchronization protocols for parallel simulation.
We have shown that there are indeed models which can be simulated significantly faster using either synchronization protocol.
\textit{Dxex} allows the user to choose between any of the six supported synchronization protocols.
We distinguish between inter-kernel (none, conservative, and optimistic synchronization) and intra-kernel (sequential or parallel concurrent transitions) synchronization protocols, which are orthogonal to one another.
Depending on observed model behaviour and simulation performance, runtime switching between synchronization protocols can be used.

Notwithstanding our modularity, \textit{dxex} achieves performance competitive to \textit{adevs}, another very efficient \textsf{DEVS} simulation tool.
Performance is measured both in elapsed time, and memory usage.
Our empirical analysis shows that allocation of models over kernels is critical to enable a parallel speedup.
Furthermore we have shown when and why optimistic synchronization can outperform conservative and vice versa.
We have also shown the influence of using the parallelism inherent in the Parallel DEVS formalism, and its interaction with inter-kernel synchronization algorithms.
Finally we investigated the effect of memory (de)allocation on parallel simulation.

Future work is possible in several directions.
First, our implementation of optimistic synchronization should be more tolerant to low-memory situations.
In its current state, simulation will simply halt with an out-of-memory error.
Having simulation control, which can throttle the speed of nodes that use up too much memory, has been shown to work in these situations~\cite{FujimotoBook}.
Faster GVT implementations~\cite{Fujimoto:1997:CGV:268403.268404,Bauer:2005:SND:1069810.1070159} might further help to minimize this problem.
Second, the runtime switching between synchronization protocols can be driven using machine learning techniques.
The simulation engine is already capable of collecting data to inform such a process, and is designed to listen for commands from an external component.
Third, automatic allocation might be possible by analysis of the connections between models.
This information is already used in \textit{dxex} to construct the dependency graph in conservative synchronization.
A graph algorithm that distributes models, while avoiding cycles, could be used to offer a parallel speedup in either optimistic or conservative synchronization.
Similarly, it could serve as a default parallel allocation scheme that can be improved by the user.
Finally, the use of transactional memory can offer several advantages in this project.
If it becomes part of the new C++17 standard it would be of great interest to see if it can help reduce the performance effects of memory allocation and synchronization.
